Some examples in extended logic counting classes

### 2017-01-05 LICS/appendix-syntactic.tex
05 ene

\subsection*{Proof of Theorem \ref{theo-pnf-snf}}
Recall that a formula in $\eqso(\LL)$ is on the following grammar:
$$
\alpha = \varphi \ \mid \ s \ \mid \ (\alpha + \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \sa{X} \alpha,
$$
where $\varphi$ is a formula in $\LL$ and $s\in\nat$. We will show a recursive function $\tau$ such that for every $\eqso(\LL)$ formula $\alpha$, $\tau(\alpha)$ is a formula in $\LL$-SNF that describes the same function as $\alpha$. If $\alpha = \varphi$ or $\alpha = s$, then we define $\tau(\alpha) = \alpha$ which clearly describes the same function, and is already in $\LL$-SNF. If $\alpha = (\alpha_1 + \alpha_2)$, then $\tau(\alpha) = (\tau(\alpha_1) + \tau(\alpha_2))$ which satisfies the condition. If $\alpha = \sa{x}\beta$, then let $\tau(\beta) = c + \sum_{i = 1}^{n}\beta_i$ where each $\beta_i$ is in $\LL$-PNF. First, if $c > 0$, we replace $c$ with the formula $(\top \add \cdots \add \top)$ (c times) and we obtain a formula $\sum_{i = 1}^{m}\beta_i$ where each $\beta_i$ is in $\LL$-PNF. If $c = 0$ the formula is already in that form. Since $\beta_i$ is in $\LL$-PNF, then $\sa{x}\beta_i$ also is, and then we define $\tau(\alpha) = \sum_{i = 1}^{m}\sa{x}\beta_i$ which satisfies the condition. If $\alpha = \sa{X}\beta$, then we proceed analogously as in the previous case. This covers all possible cases for $\alpha$ and we conclude the proof by taking $\tau(\alpha)$ as the desired rewrite of $\alpha$.

@@

\vspace{1em}
We now show that if $\LL$ contains $\logu{1}$ and is closed under conjunction and disjunction, then for every formula $\alpha$ in $\eqso(\LL)$ there is an equivalent formula $\beta$ in $\LL$-PNF. As in Theorem \ref{theo-pnf-snf}, we show a recursive function $\tau$ that produces such formula. As we showed, there exists an equivalent formula in $\LL$-SNF, so we assume that $\alpha$ is in that form. Let $\alpha = c + \sum_{i = 1}^n \alpha_i$ where each $\alpha_i$ is in $\LL$-SNF. First, as we did in Theorem \ref{theo-pnf-snf}, we replace the constant $c$ for $(\top \add \cdots \add \top)$ if $c > 0$ and we do nothing if $c = 0$, to obtain an equivalent formula $\sum_{i = 1}^m \alpha_i$. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi_i(\bar{X},\bar{x})$. We replace each $\alpha_i$ for the equivalent formula $\sa{\bar{X}}\sa{Y}\sa{\bar{x}}\sa{y}(\varphi_i(\bar{X},\bar{x})\wedge\forall z\,Y(z) \wedge \forall z(y \leq z))$ to obtain that $\length(\bar{X}) > 0$ and $\length(\bar{x}) > 0$.

@@

We suppose that $\alpha$ is in SNF, this is, $\alpha = c + \sum_{i = 1}^n\alpha_i$. Let $\mathbb{1}$ be a $\R$-structure with one element. Note that if $c > 0$, then $\alpha(\enc(\mathbb{1})) \geq c > 0$ which is not possible since $\alpha(\enc(\mathbb{1})) = 0$. Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$, where $\varphi$ is quantifier-free. Note that if $\length(\bar{X}) > 0$, then the function $\alpha$ describes is in $\Omega(2^n)$, as it was proven by the authors in \cite{SalujaST95}. We have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$, as we proved in the claim, describes either some function greater or equal to $n$, or in $\Omega(n^2)$, which leads to a contradiction. Lastly, note that the formula $\sa{x}\exists y(x < y)$ is in $\E{0}$ and describes the function $n-1$, which concludes the proof.

@@

We separate this proof in two parts

@@

\subsection*{Proof of Theorem \ref{theo:binary-prod}}

Let $\LL$ be a fragment of $\so$ that is closed under conjunction. We will define a recursive function $\tau$ that receives a formula $\alpha$ over the grammar of $\eqso(\LL)$ with multiplication, and outputs an equivalent formula $\tau(\alpha)$ over the unextended grammar of $\eqso(\LL)$. Furthermore, the formula $\tau(\alpha)$ is in $\LL$-SNF. First we replace each constant $s$ in $\alpha$ for $(\top \add \cdot \add \top)$ ($s$ times). If $\alpha = \varphi$, then we define $\tau(\alpha) = \alpha$. We assume that for every $\beta$ that has less algebraic operators than $\alpha$, $\tau(\beta)$ is in $\LL$-SNF. If $\alpha = (\alpha_1 + \alpha_2)$ then we define $\tau(\alpha) = \tau(\alpha_1) + \tau(\alpha_2)$. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$, then we define $\tau(\alpha)$ as the formula in $\LL$-SNF that is equivalent to $\sa{x}\tau(\beta)$ and to $\tau(\alpha) = \sa{X}\tau(\beta)$, respectively. If $\alpha = (\alpha_1 \cdot \alpha_2)$, we assume that each $\alpha_i$ is in $\LL$-SNF. We identify three cases. (1) Some $\alpha_i$ is equal to $\sum_{j = 1}^n\beta_j$ for $n > 1$. Suppose wlog. that it is $\alpha_1$. We then define $\tau(\alpha) = \sum_{j = 1}^n\tau(\beta_j\cdot\alpha_2)$. In the following cases, $\alpha_1$ and $\alpha_2$ are in $\LL$-SNF. (2) If some $\alpha_i$ is equal to $\sa{X}\beta$ or $\sa{x}\beta$, we define $\tau(\alpha)$ as the $\LL$-SNF formula that is equivalent to $\sa{x}\tau(\beta\cdot\alpha_2)$ and $\sa{X}\tau(\beta\cdot\alpha_2)$, respectively. The remaining case is (3) $\alpha_1 = \varphi_1$ and $\alpha_2 = \varphi_2$ where each $\varphi$ is an $\LL$ formula. Then we define $\tau(\alpha) = \varphi_1 \wedge \varphi_2$. This covers all possible cases for $\alpha$. For every formula $\alpha$ in $\eqso(\LL)$ extended by multiplication, we take $\tau(\alpha)$ which is equivalent and in unextended $\eqso(\LL)$. This concludes the proof.

@@

\subsection*{Proof of Proposition \ref{prop:uhorn-pe}} %V.11
This can be obtained directly by Proposition \ref{prop:ehorn-pe}, since $\eqso(\uhorn)\subseteq\eqso(\ehorn)\subseteq\totp$.

>>>>

\subsection*{Proof of Theorem \ref{theo-pnf-snf}}
Recall that a formula in $\eqso(\LL)$ is on the following grammar:
$$
\alpha = \varphi \ \mid \ s \ \mid \ (\alpha + \alpha) \ \mid \ \sa{x} \alpha \ \mid \ \sa{X} \alpha,
$$
where $\varphi$ is a formula in $\LL$ and $s\in\nat$. We will construct a recursive function $\tau$ such that for every $\eqso(\LL)$ formula $\alpha$, it outputs an equivalent formula $\tau(\alpha)$ which is in $\LL$-SNF. If $\alpha = \varphi$, then we define $\tau(\alpha) = \alpha$ which is clearly equivalent and already in $\LL$-SNF. If $\alpha = s$, then we define $(\top \add \cdots \add \top)$ ($s$ times), which also satisfies the condition. We assume that for every sub-formula $\beta$ of $\alpha$, $\tau(\beta)$ is an equivalent formula in $\LL$-SNF. If $\alpha = (\alpha_1 + \alpha_2)$, then we define $\tau(\alpha) = (\tau(\alpha_1) + \tau(\alpha_2))$. If $\alpha = \sa{x}\beta$, then let $\tau(\beta) = \sum_{i = 1}^{n}\beta_i$ where each $\beta_i$ is in $\LL$-PNF. We have that each $\beta_i$ also is in $\LL$-PNF, and then we define $\tau(\alpha) = \sum_{i = 1}^{m}\sa{x}\beta_i$ which satisfies the condition. If $\alpha = \sa{X}\beta$, then we proceed analogously as in the previous case. This covers all possible cases for $\alpha$ and we conclude the proof by taking $\tau(\alpha)$ as the desired rewrite of $\alpha$.

@@

\vspace{1em}
We now show that if $\LL$ contains $\logu{1}$ and is closed under conjunction and disjunction, then for every formula $\alpha$ in $\eqso(\LL)$ there is an equivalent formula $\beta$ in $\LL$-PNF. As in Theorem \ref{theo-pnf-snf}, we show a recursive function $\tau$ that produces such formula. As we showed, there exists an equivalent formula in $\LL$-SNF, so we assume that $\alpha$ is in that form. Let $\alpha = \sum_{i = 1}^n \alpha_i$ where each $\alpha_i$ is in $\LL$-SNF. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi_i(\bar{X},\bar{x})$. We replace each $\alpha_i$ for the equivalent formula $\sa{\bar{X}}\sa{Y}\sa{\bar{x}}\sa{y}(\varphi_i(\bar{X},\bar{x})\wedge\forall z\,Y(z) \wedge \forall z(y \leq z))$ to obtain that $\length(\bar{X}) > 0$ and $\length(\bar{x}) > 0$.

@@

We suppose that $\alpha$ is in SNF, this is, $\alpha = \sum_{i = 1}^n\alpha_i$. Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$, where $\varphi$ is quantifier-free. Note that if $\length(\bar{X}) > 0$, then the function $\alpha$ describes is in $\Omega(2^n)$, as it was proven by the authors in \cite{SalujaST95}. We have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$, as we proved in the claim, describes either some function greater or equal to $n$, or in $\Omega(n^2)$, which leads to a contradiction. Lastly, note that the formula $\sa{x}\exists y(x < y)$ is in $\E{0}$ and describes the function $n-1$, which concludes the proof.

@@

We separate this proof in three parts

@@

\vspace{2em}
For the third part, we will prove that $\eqso(\loge{1})$ is closed under sum and multiplication. Since $\eqso(\loge{1})$ is closed under sum by definition, we focus only in proving that the class is closed under multiplication. We prove this for a more general case for $\eqso(\LL)$ where $\LL$ is a fragment of $\so$.

Let $\LL$ be a fragment {\bf closed under conjunction}. We will define a recursive function $\tau$ that receives a formula $\alpha$ over the grammar of $\eqso(\LL)$ extended by binary product, and outputs an equivalent formula $\tau(\alpha)$ over the unextended grammar of $\eqso(\LL)$. In fact, the formula $\tau(\alpha)$ is in $\LL$-SNF. First we replace each constant $s$ in $\alpha$ for $(\top \add \cdot \add \top)$ ($s$ times). If $\alpha = \varphi$, then we define $\tau(\alpha) = \alpha$. We assume that for every $\beta$ that has less algebraic operators than $\alpha$, $\tau(\beta)$ is in $\LL$-SNF. If $\alpha = (\alpha_1 + \alpha_2)$ then we define $\tau(\alpha) = \tau(\alpha_1) + \tau(\alpha_2)$. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$, then we define $\tau(\alpha)$ as the formula in $\LL$-SNF that is equivalent to $\sa{x}\tau(\beta)$ and to $\tau(\alpha) = \sa{X}\tau(\beta)$, respectively. If $\alpha = (\alpha_1 \cdot \alpha_2)$, we assume that each $\alpha_i$ is in $\LL$-SNF. We identify three cases. (1) Some $\alpha_i$ is equal to $\sum_{j = 1}^n\beta_j$ for $n > 1$. Suppose wlog. that it is $\alpha_1$. We then define $\tau(\alpha) = \sum_{j = 1}^n\tau(\beta_j\cdot\alpha_2)$. In the following cases, $\alpha_1$ and $\alpha_2$ are in $\LL$-SNF. (2) If some $\alpha_i$ is equal to $\sa{X}\beta$ or $\sa{x}\beta$, we define $\tau(\alpha)$ as the $\LL$-SNF formula that is equivalent to $\sa{x}\tau(\beta\cdot\alpha_2)$ and $\sa{X}\tau(\beta\cdot\alpha_2)$, respectively. The remaining case is (3) $\alpha_1 = \varphi_1$ and $\alpha_2 = \varphi_2$ where each $\varphi$ is an $\LL$ formula. Then we define $\tau(\alpha) = \varphi_1 \wedge \varphi_2$. This covers all possible cases for $\alpha$.

For every pair of formulas $\alpha,\beta$ in $\eqso(\LL)$, we have that their multiplication $(\alpha\cdot\beta)$ is a formula in the grammar $\eqso(\LL)$ extended by binary product, and so, there exists an equivalent formula $\tau(\alpha\cdot\beta)$ which is in unextended $\eqso(\LL)$. Since $\loge{1}$ is closed under conjunction, this also holds for $\eqso(\loge{1})$. This concludes the proof.

@@
